---
title: "Lab_4_assignment_group_09"
author: 
  - "Andreis Marco(s243116)" 
  - "Asiya Mohamad Yusuf Muse (s243930)"
  - "Magnus Harthimmer (s233426)"  
  - "Anita Skovbjerg Hjort-Gregersen (s201190)"
  - "Rebecca Hjermind Millum (s215024)"
format: html
editor: visual
---

For this assignment we are going to follow [this](https://clauswilke.com/blog/2020/09/07/pca-tidyverse-style/) blog post, which will serve as an toy example to test the possibilities of collaboration provided by Git. In particular each of the four step described in the blog will be done by a different member of the group.

```{r}
# import necessary packages
suppressMessages(library(tidyverse))
suppressMessages(library(broom))
suppressMessages(library(cowplot))
```

The dataset we are going to be working with consist in biopsies of breast cancer patients, for which have been collected 9 discrete attributes (1-10) and an outcome classifying the tumors either as malignant or benign. With this analysis we want to get an idea of how the different attributes contribute to determining the outcome.

```{r}
# import dataset
biopsy <- read_csv("https://wilkelab.org/classes/SDS348/data_sets/biopsy.csv",
                   show_col_types = FALSE)
```

## Look at the data in PC coordinates

```{r}
pca_fit <- biopsy %>% 
  select(where(is.numeric)) %>% # retain only numeric columns
  prcomp(scale = TRUE) # do PCA on scaled data
```

```{r}
pca_fit <- biopsy %>% 
  select(where(is.numeric)) %>% # retain only numeric columns
  scale() %>% # scale data
  prcomp() # do PCA
```

```{r}
pca_fit %>%
  augment(biopsy) %>% # add original dataset back in
  ggplot(aes(.fittedPC1, .fittedPC2, color = outcome)) + 
  geom_point(size = 1.5) +
  scale_color_manual(
    values = c(malignant = "#D55E00", benign = "#0072B2")
  ) +
  theme_half_open(12) +
  background_grid()
```

From the plot we can see that the malignant are much more diverse than the benign ones, which instead cluster more nicely.

## Look at the data in PC coordinates

```{r}
# extract rotation matrix
pca_fit %>%
  tidy(matrix = "rotation")
```

```{r}
# define arrow style for plotting
arrow_style <- arrow(
  angle = 20, ends = "first", type = "closed", length = grid::unit(8, "pt")
)

# plot rotation matrix
pca_fit %>%
  tidy(matrix = "rotation") %>%
  pivot_wider(names_from = "PC", names_prefix = "PC", values_from = "value") %>%
  ggplot(aes(PC1, PC2)) +
  geom_segment(xend = 0, yend = 0, arrow = arrow_style) +
  geom_text(
    aes(label = column),
    hjust = 1, nudge_x = -0.02, 
    color = "#904C2F"
  ) +
  xlim(-1.25, .5) +
  ylim(-.5, 1) +
  coord_fixed() + # fix aspect ratio to 1:1
  theme_minimal_grid(12)
```

We can see that 8 of the 9 attributes mainly contribute to PC1, while the main contributer to PC2 is the "mitosis" attribute.

## Look at the variance explained by each PC

```{r}
pca_fit %>%
  tidy(matrix = "eigenvalues")
```

```{r}
pca_fit %>%
  tidy(matrix = "eigenvalues") %>%
  ggplot(aes(PC, percent)) +
  geom_col(fill = "#56B4E9", alpha = 0.8) +
  scale_x_continuous(breaks = 1:9) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.01))
  ) +
  theme_minimal_hgrid(12)
```

We can see that PC1 is capable of explaning more than 65% of the total variance of the datapoints, which is compatible with the first PCA plot.
